Reconstructing the Lensing Mass in the Universe \\ from Photometric Catalog Data

### 2012-10-31 doc/paper-lcr.tex
abstract done - shipping to collaborators!

Cosmological dark matter simulations have shown that dark matter
halos are well-approximated by NFW profiles \citep{NFW}, with a density
profile given by
\be\label{eq:rhonfw}
\rho(r) = 
\frac{\rho_0}{\left(r/r_{s}\right)
\left(1+r/r_{s}\right)^{2}},
\ee
where $r_{s}$ is a characteristic radius of the cluster, representing where
the density slope transitions from $r^{-1}$ to $r^{-3}$. This radius is related
to the virial radius of the halo by $r_{s}~=~r_{200}/c$, where $r_{200}$ is the radius at 
which 
\be
\rho(r_{200}) = 200\rho_{{\mathrm{crit}}}(z), \text{ where } \rho_{\mathrm{crit}}(z) \equiv \frac{3H^{2}(z)}{8\pi G}
\ee
and $c$ is the concentration parameter, which can be estimated from the halo's mass,
using a mass--concentration relation. Typically more massive halos are less concentrated,
but there is some scatter; we use the relation of \citet{Neto2007} to estimate $c$
from the mass enclosed within $r_{200}$, which we denote as $M_{200}$. We find our results do not change if the \citet{MaccioEtal2008} Mass--concencetration relation is used instead. The best fit relation of \citet{Neto2007} is given by
\be
c_{200} = 4.67 (M_{200}/10^{14} h^{-1} \Msun)^{-0.11}.
\ee
\todo{Tom}{Add sentence about scatter!}

@@

This PDF contains a hidden assumption of an uninformative prior PDF for
$\kappah$ -- in the case of infinitely poorly measured stellar masses and
redshifts, $\pr(\kappah|\data)$ will have very long tails corresponding to
very overdense sightlines that we do not believe exist. The resolution of this
is to divide out the effective prior PDF that was applied during the halo mass
estimation process (which is broad, and hance close to uniform), and apply an
additional prior on $\kappah$, given by the global underlying $\kappah$
distribution from the simulation. In the limiting  case of very poor
photometric data, $\pr(\kappah|\data)$ then defaults to this distribution.
\todo{All}{Discuss this. It ties us to the simulation, but only weakly. Is the
argument about poor data valid? Tom?}

@@

The solution we present here is to calibrate the relationship between halo
convergence $\kappah$ and underlying convergence $\kappa$ (as calculated
during full ray tracing) using the simulations themselves.
The halo modelling \proceedure outlined in \Sref{sec:model:halos} allows 
us to estimate 
the convergence due to halos given observations of 
mass, redshift and position,
$\pr(\kappah|\mathcal{D})$, but what we are interested in is the total convergence
along the line of sight $\pr(\kappa|\mathcal{D})$. We can obtain this by
considering the expression:
\begin{equation}
\Pr(\kappa|\mathcal{D}) = \int \dee\kappah 
   \Pr(\kappa|\kappah) \Pr(\kappah|\mathcal{D})
\label{eq:kappaconv}   
\end{equation}
The first term in the integrand relates the convergence due to model halos,
$\kappah$, to the true convergence, $\kappa$. This conditional distribution
can be constructed from the Millennium Simulation catalogs, by computing
$\kappah$ for each selected line of sight using the true values of the halo
parameters,  and accumulating $\kappah$ and $\kappa$ along a large number of
such lines of
sight to  characterize $\Pr(\kappa|\kappah)$.
For any given line of sight we can then estimate
$\pr(\kappah|\mathcal{D})$ as described above, and then
multiply it by $\Pr(\kappa|\kappah)$ and integrate out the intermediate
nuisance parameter $\kappah$. 
% We can do the marginalisation over $\kappah$
% by drawing samples from the two-dimensional grid and keeping only the
% $\kappa$ values, to form our final result, $\Pr(\kappa|\mathcal{D})$.

@@

We find that we can correct this bias to some extent by artificially 
suppressing the tails of $\pr(\kappah|\{\Mstar,z\})$. Let the median of
this PDF be denoted $\kappah^{\rm med}$. We again take a large number of
calibration sightlines from the \MS catalogs, but this time generate a mock
stellar mass catalog for each one. We then reconstruct each lightcone in the
same way as we would an observed field,
estimate $\pr(\kappah|\{\Mstar,z\})$, and extract its median. Accumulating
these samples and their corresponding true values of $kappa$, we can form the 
conditional distribution $\pr(\kappa|\kappah^{\rm med})$. 
Then, when inferring $\kappa$ from an observed photometric catalog, we would
estimate $\pr(\kappah|\{\Mstar,z\}_{\rm obs})$ 
using the halo model, compute the median value 
$\kappa_{\rm h,obs}^{\rm med}$, and then use the following alternative
version of 
\Eref{eq:kappaconv} to infer $\kappa$: 
\bea
\Pr(\kappa|\mathcal{D}) &=& \int \dee\kappah^{\rm med} 
   \Pr(\kappa|\kappah^{\rm med}) \Pr(\kappah^{\rm med}|\mathcal{D}) \notag \\
                        &=& \Pr(\kappa|\kappa_{\rm h,obs}^{\rm med}).
\label{eq:calkappaconv}   
\eea
Here, we have replaced $\Pr(\kappah^{\rm med}|\mathcal{D})$ with a delta
function centred at the observed median value $\kappa_{\rm h,obs}$.

@@

\Fref{fig:where} shows the cumulative contribution to $\kappah$ as a function
of projected distance from the line of sight, magnitude and redshift.
We find that most of the convergence comes from halos whose
galaxies are close to the line of sight. 
Over half of the  convergence
typically comes from halos within 30 arcseconds of the line of sight; by 2
arcminutes, this fraction is XX\%. \todo{Tom}{Put a number on this.} 
As a function of magnitude, we find
that $\kappah$ is dominated by objects with magnitudes between $i=18$ and
$i=24$. Objects brighter than $i=18$ are either too rare, or too close to the
observer, to make a significant contribution to the convergence. Objects
fainter than $i=24$ are too small to be important, unless they are extremely
close to the line of sight, but we find that including halos fainter than
$i=24$ does not improve the reconstruction.  \Fref{fig:magcut} shows the
scatter on $\kappah-\kappa$ (where $\kappah$ is shifted so that its mean is
zero) as a function of reconstruction magnitude limit. 
%
% Since these objects
% must be extremely close to the line of sight to make a non-zero $\kappah$
% contribution, it is likely that neglecting stellar mass and using a spherical
% NFW prescription are too naive to adequately reconstruct $\kappax$.
%
% PJM: the sentence above does not make sense!
%
We do
not expect a deeper survey to decrease the size of the uncertainty in mapping
$\kappah$ onto $\kappa$. Halos at all redshifts out to the source redshift
(1.4, here) contribute to the convergence, but the largest contribution comes from
halos with $z \sim z_{\rm source}/2$.

@@

While the \MS catalogs do already contain stellar masses for each galaxy, we
find that these are not well behaved... \todo{Tom}{Explain why we re-made the
stellar masses}. We also wanted to be able to perform the functional test of
trying to recover the convergence having assumed the correct stellar
mass--halo mass relation. For these reasons, we assign a new stellar mass to
each halo in the \MS catalogs, according to the stellar mass--halo mass
relation of  \citet{BehrooziEtal2010}. From this true  stellar mass we
simulate an observed stellar mass by drawing a sample from 
$\pr(\log{M^{*}_{\rm obs}}|\log{M^{*}_{\rm true}})$ which we take to be a
Gaussian of width $\sigma_{M_*}$ centred on $\log(M*_{\mathrm {true}})$. Where
a spectroscopic redshift exists stellar masses can be estimated with typical
uncertainties of 0.15 dex \citep{AugerEtal2009}; however with photometric
redshifts stellar mass uncertainties are typically three times as large. We
use $\sigma_{M_*}=0.15$ for halos with a spectroscopic redshift and
$\sigma_{M_*}=0.45$ otherwise. For photometric redshift errors we draw a
redshift from $\pr(z_{\rm true}|z_{\rm obs})$ which we take to be a Gaussian
of width $0.1(1+z_{\rm spec})$ centred on $z_{\rm spec}$, where $z_{\rm spec}$
is the halo's true redshift in the \MS catalogue.

@@

\begin{figure}
\includegraphics[width=\columnwidth]{figs/Widthvshilberrt.eps}
\caption{Width of the \infered PDF $\pr(\kappah|\data)$ versus the true
$\kappa$ along a line of sight. Blue assumes a spectroscopic redshift for
every halo, whilst green assumes only photometric redshifts. The width of the
\infered $\pr(\kappah|\data)$ increases with $\kappa$. 
\todo{Tom}{Check that you mean
$\kappah$ and not $\kappa$, and potentially change the graph!}}
\label{fig:widthsvsH}
\end{figure}

>>>>

Cosmological dark matter simulations have shown that dark matter
halos are well-approximated by NFW profiles \citep{NFW}, with a density
profile given by
\be\label{eq:rhonfw}
\rho(r) = 
\frac{\rho_0}{\left(r/r_{s}\right)
\left(1+r/r_{s}\right)^{2}},
\ee
where $r_{s}$ is a characteristic radius of the cluster, representing where
the density slope transitions from $r^{-1}$ to $r^{-3}$. This radius is related
to the virial radius of the halo by $r_{s}~=~r_{200}/c$, where $r_{200}$ is the radius at 
which 
\be
\rho(r_{200}) = 200\rho_{{\mathrm{crit}}}(z), \text{ where } \rho_{\mathrm{crit}}(z) \equiv \frac{3H^{2}(z)}{8\pi G}
\ee
and $c$ is the concentration parameter, which can be estimated from the halo's mass,
using a mass--concentration relation. Typically more massive halos are less concentrated,
but there is some scatter; we use the relation of \citet{Neto2007} to estimate $c$
from the mass enclosed within $r_{200}$, which we denote as $M_{200}$. We find our results do not change if the \citet{MaccioEtal2008} Mass--concencetration relation is used instead. The best fit relation of \citet{Neto2007} is given by
\be
c_{200} = 4.67 (M_{200}/10^{14} h^{-1} \Msun)^{-0.11}.
\ee

@@

This PDF contains a hidden assumption of an uninformative prior PDF for
$\kappah$ -- in the case of infinitely poorly measured stellar masses and
redshifts, $\pr(\kappah|\data)$ will have very long tails corresponding to
very overdense sightlines that we do not believe exist. The resolution of this
is to divide out the effective prior PDF that was applied during the halo mass
estimation process (which is broad, and hance close to uniform), and apply an
additional prior on $\kappah$, given by the global underlying $\kappah$
distribution from the simulation. In the limiting  case of very poor
photometric data, $\pr(\kappah|\data)$ then defaults to this distribution.

% \todo{All}{Discuss this. It ties us to the simulation, but only weakly. Is the
% argument about poor data valid? Tom?}

@@

The solution we present here is to calibrate the relationship between halo
convergence $\kappah$ and underlying convergence $\kappa$ (as calculated
during full ray tracing) using the simulations themselves.
The halo modelling \proceedure outlined in \Sref{sec:model:halos} allows 
us to estimate 
the convergence due to halos given observations of 
mass, redshift and position,
$\pr(\kappah|\mathcal{D})$, but what we are interested in is the total convergence
along the line of sight $\pr(\kappa|\mathcal{D})$. We can obtain this by
considering the expression:
\begin{equation}
\Pr(\kappa|\mathcal{D}) = \int \dee\kappah 
   \Pr(\kappa|\kappah) \Pr(\kappah|\mathcal{D})
\label{eq:kappaconv}   
\end{equation}
The first term in the integrand relates the convergence due to model halos,
$\kappah$, to the true convergence, $\kappa$. This conditional distribution
can be constructed from the Millennium Simulation catalogs, by computing
$\kappah$ for each selected line of sight using the true values of the halo
parameters,  and accumulating $\kappah$ and $\kappa$ along a large number of
such lines of
sight to  characterize $\Pr(\kappa|\kappah)$.
For any given line of sight we can then estimate
$\pr(\kappah|\data)$ as described above, and then
multiply it by $\Pr(\kappa|\kappah)$ and integrate out the intermediate
nuisance parameter $\kappah$. 
% We can do the marginalisation over $\kappah$
% by drawing samples from the two-dimensional grid and keeping only the
% $\kappa$ values, to form our final result, $\Pr(\kappa|\mathcal{D})$.

@@

We find that we can correct this bias to some extent by artificially 
suppressing the tails of $\pr(\kappah|\{\Mstar,z\})$. Let the median of
this PDF be denoted $\kappah^{\rm med}$. We again take a large number of
calibration sightlines from the \MS catalogs, but this time generate a mock
stellar mass catalog for each one. We then reconstruct each lightcone in the
same way as we would an observed field,
estimate $\pr(\kappah|\{\Mstar,z\})$, and extract its median. Accumulating
these samples and their corresponding true values of $\kappa$, we can form the 
conditional distribution $\pr(\kappa|\kappah^{\rm med})$. 
Then, when inferring $\kappa$ from an observed photometric catalog, we would
estimate $\pr(\kappah|\{\Mstar,z\}_{\rm obs})$ 
using the halo model, compute the median value 
$\kappa_{\rm h,obs}^{\rm med}$, and then use the following alternative
version of 
\Eref{eq:kappaconv} to infer $\kappa$: 
\bea
\Pr(\kappa|\mathcal{D}) &=& \int \dee\kappah^{\rm med} 
   \Pr(\kappa|\kappah^{\rm med}) \Pr(\kappah^{\rm med}|\mathcal{D}) \notag \\
                        &=& \Pr(\kappa|\kappa_{\rm h,obs}^{\rm med}).
\label{eq:calkappaconv}   
\eea
Here, we have replaced $\Pr(\kappah^{\rm med}|\mathcal{D})$ with a delta
function centred at the observed median value $\kappa_{\rm h,obs}$. 
The procedure of reducing the full PDF for $\kappah$ to just its median is
risky: the simulated lightcones that have median $\kappah$ values to match are
not guaranteed to provide an accurate estimate of $\kappa$ via this PDF
look-up. We proceed with caution, quantifying the biases and PDF widths in the
sections below.

@@

\Fref{fig:where} shows the cumulative contribution to $\kappah$ as a function
of projected distance from the line of sight, magnitude and redshift.
We find that most of the convergence comes from halos whose
galaxies are close to the line of sight. 
Over half of the  convergence
typically comes from halos within 30 arcseconds of the line of sight; by 2
arcminutes, this fraction is \todo{XX\%}{Tom?}.
As a function of magnitude, we find
that $\kappah$ is dominated by objects with magnitudes between $i=18$ and
$i=24$. Objects brighter than $i=18$ are either too rare, or too close to the
observer, to make a significant contribution to the convergence. Objects
fainter than $i=24$ are too small to be important, unless they are extremely
close to the line of sight, but we find that including halos fainter than
$i=24$ does not improve the reconstruction.  \Fref{fig:magcut} shows the
scatter on $\kappah-\kappa$ (where $\kappah$ is shifted so that its mean is
zero) as a function of reconstruction magnitude limit. 
%
% Since these objects
% must be extremely close to the line of sight to make a non-zero $\kappah$
% contribution, it is likely that neglecting stellar mass and using a spherical
% NFW prescription are too naive to adequately reconstruct $\kappax$.
%
% PJM: the sentence above does not make sense!
%
We do
not expect a deeper survey to decrease the size of the uncertainty in mapping
$\kappah$ onto $\kappa$. Halos at all redshifts out to the source redshift
(1.4, here) contribute to the convergence, but the largest contribution comes from
halos with $z \sim z_{\rm source}/2$.

@@

While the \MS catalogs do already contain stellar masses for each galaxy, we
find that these are not well behaved at the low mass end. 
% \todo{Tom}{Explain why we re-made the stellar masses!}. 
We also wanted to be able to perform the functional test of
trying to recover the convergence having assumed the correct stellar
mass--halo mass relation. For these reasons, we assign a new stellar mass to
each halo in the \MS catalogs, according to the stellar mass--halo mass
relation of  \citet{BehrooziEtal2010}. From this true  stellar mass we
simulate an observed stellar mass by drawing a sample from 
$\pr(\log{M^{*}_{\rm obs}}|\log{M^{*}_{\rm true}})$ which we take to be a
Gaussian of width $\sigma_{M_*}$ centred on $\log(M*_{\mathrm {true}})$. Where
a spectroscopic redshift exists stellar masses can be estimated with typical
uncertainties of 0.15 dex \citep{AugerEtal2009}; however with photometric
redshifts stellar mass uncertainties are typically three times as large. We
use $\sigma_{M_*}=0.15$ for halos with a spectroscopic redshift and
$\sigma_{M_*}=0.45$ otherwise. For photometric redshift errors we draw a
redshift from $\pr(z_{\rm true}|z_{\rm obs})$ which we take to be a Gaussian
of width $0.1(1+z_{\rm spec})$ centred on $z_{\rm spec}$, where $z_{\rm spec}$
is the halo's true redshift in the \MS catalogue.

@@

\begin{figure}
\includegraphics[width=\columnwidth]{figs/Widthvshilberrt.eps}
\caption{Width of the \infered PDF $\pr(\kappah|\data)$ versus the true
$\kappa$ along a line of sight. Blue assumes a spectroscopic redshift for
every halo, whilst green assumes only photometric redshifts. The width of the
\infered $\pr(\kappah|\data)$ increases with $\kappa$.}
%\todo{Tom}{Check that you mean $\kappah$ and not $\kappa$, and potentially change the graph!}}
\label{fig:widthsvsH}
\end{figure}
